What We Eval in the Shadows

### 2021-08-11 paper/main.tex
sec 6 clean up

The R language was intended to be extensible. The combination of lazy
evaluation, \c{substitute} and \c{eval} are some of the tools given to
developers to this end. To understand \emph{why} developers use \eval, we
manually inspected top 117 \eval call-sites which contribute to 90\% of the
\eval calls. In this section, we give examples of the \eval usage from this
real-world sample. The examples are often simplified to fit the space and
increase clarity. For each example, we indicate the package and the function
where it is located.

\paragraph{Simplify Interfaces} A common use of \eval is to simplify interfaces by
letting users pass string argument which can be parsed and evaluated in custom a
environment. An example is the \code{rARS} function shown below. This function
is defined in the \code{AdapSamp} package that implements sampling algorithms.

\begin{figure}[h]
\begin{lstlisting}
rARS <- function(n, formula, min=-Inf, max=Inf, sp) {
    ...
    p <- function(x) { eval(parse(text=formula)) }
    ...
}
\end{lstlisting}
  \caption{AdapSamp::rARS}
\end{figure}

The \code{rARS} function take a \code{formula} argument, an expression in
\code{x} as a string, that represents the kernel of the target distribution.
This is turned into the closure \code{function(x) \{ eval(parse(text=formula))
  \}}. An alternative design would be to supply the closure itself as the
argument, but passing \code{function(x) exp(-x^2/2)} is probably not as
convenient for domain-experts as passing \code{"exp(-x^2/2)"}.

\paragraph{Code Generation} The \c{Rcpp} package which bridges R with C++ is a
popular user of this pattern. Its \code{.makeCppMethods} function, shown below,
uses \eval to automatically generate R bindings to native code. The expression
\code{function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))} is
evaluated with different values of \code{what} to generate wrapper functions
which are assigned to the \code{methods} structure keyed by \code{what}.

\begin{figure}[h]
\begin{lstlisting}
.makeCppMethods <- function(methods, cppclassinfo, env) {
    ...
    for(what in cppMethods[! cppMethods %in% newMethods])
        methods[[what]] <- eval(substitute(
                  function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))), env)
    ...
}
\end{lstlisting}
  \caption{Rcpp::.makeCppMethods}
\end{figure}

\paragraph{Extracting varargs} A surprising use of \eval is the
extraction of unevaluated text of arguments passed to a \code{...} parameter.
Consider the definition of the \code{NVL} function from the
\code{statnet.common} package. This function returns the first argument that is
non \code{NULL}.

\begin{figure}[h]
\begin{lstlisting}
NVL <- function(...) {
    for(e in eval(substitute(alist(...))))
        x <- eval(e, parent.frame())
        if(!is.null(x)) break
    }
    x
}
\end{lstlisting}
  \caption{statnet.common::NVL}
\end{figure}

The expression \code{eval(substitute(alist(...)))} is used to extract the
unevaluated text of the arguments passed to \code{NVL}.
\code{substitute(alist(...))} returns an expression representing a call to the
\code{alist} function with \code{...} replaced by the arguments supplied to
\code{NVL}. Evaluating this expression returns a \code{list} whose arguments are
the unevaluated argument expressions which are evaluated sequentially for their
result.

\paragraph{Conditional Evaluation} Developers frequently use \eval to explicitly
control the evaluation of code. Consider the \code{\%\|\|\%} operator from the
\code{igraph} package shown below. This operator returns the result of
evaluating the first argument (\code{lhs}) if it is not \code{NULL}, otherwise
it returns the result of evaluating the second argument (\code{rhs}). This is
similar to the \code{\|\|} operator which performs the logical OR of its
arguments except it treats \code{NULL} as the \code{FALSE} value.

\begin{figure}[h]
\begin{lstlisting}
`%||%` <- function (lhs, rhs) {
    lres <- withVisible(eval(lhs, envir = parent.frame()))
    if (is.null(lres$value)) {
        eval(rhs, envir = parent.frame())
    } else {
        if (lres$visible) { lres$value}
        else { invisible(lres$value) }
   }
}
\end{lstlisting}
  \caption{igraph::\%||\%}
\end{figure}

A similar idea is used by assertion packages such as \code{assertthat} which
provide functions to check pre and post conditions and produce friendly error
messages on failures. These conditions are written as expressions which are
evaluated using \eval. Consider the \code{see\_if} function shown below. It
accepts the conditions as arguments to the \code{...} parameter. The code of
these conditions is extracted using \code{eval(substitute(alist(...)))}. The
conditions are then evaluated sequentially by the expression
\code{eval(assertion, env)}, inside a \code{tryCatch} block. If a condition
evaluates to \code{FALSE}, the function returns immediately with an error
message, leaving the remaining conditions unevaluated.

\begin{figure}[h]
\begin{lstlisting}
see_if <- function(..., env = parent.frame(), msg = NULL) {
    asserts <- eval(substitute(alist(...)))
    for (assertion in asserts) {
      res <- tryCatch({
        eval(assertion, env)
      }, assertError = function(e) {
        structure(FALSE, msg = e$message)
      })
    ...
}
\end{lstlisting}
  \caption{assertthat::see\_if}
\end{figure}

\paragraph{Core function variants} Many packages use \eval to implement variations of
the
\code{base} R functions. For example, the \code{igraph} package provides the
\code{do_call} function shown below. This is a variant of the \code{do.call}
function from the \code{base} package. This function executes a function call by
constructs it from a name and a list of arguments and calling \eval.

\begin{figure}[h]
\begin{lstlisting}
do_call <- function(f, ..., .args = list(), .env = parent.frame()) {
  f <- substitute(f)
  call <- make_call(f, ..., .args)
  eval(call, .env)
}
\end{lstlisting}
\caption{igraph::do\_call}
\end{figure}

\paragraph{Domain-Specific Languages} The combination of laziness, metaprogramming,
first-class environments and dynamic evaluation makes R the perfect toolkit for
building DSLs. An example of this is the \code{glue} package which provides
functions for string interpolation. The code snippet below shows an example of
\code{glue} function in action.

@@

The \code{glue} function extracts snippets of R code enclosed between \c{\{} and
\c{\}}, evaluates them using \eval, and splices the result to construct the
final string.

Another example is the \code{data.table} package that defines a DSL for data
frame manipulation using the subset operator \code{[}. This subset operation is
defined as \code{DT[i, j, k]}, where expression \code{j} is used to calculate on
a subset/reorder of rows from \code{i} grouped by \code{k}. This enables compact
queries on data frames such as \code{flights[carrier == "AA", .N, by = origin]}
which will compute the number of flights done by American Airlines (\code{"AA"})
for each origin airport. At its core, \code{[} uses \code{substitute} to capture
the operand expressions, transforms them, and evaluates them using \eval in an
environment which has data frame columns available as bindings.

The use of \eval with \code{substitute} is a general pattern that we often see
in R packages. It is used to make the interfaces more flexible. For example the
\code{subset}, which selects columns in a data frame, uses it to allow to specify
the selected columns using either the column name or index. For example
\code{subset(df, sel=c(1,B,x))} returns a subset of \code{df} with the first
column, the column named \code{B} and whatever column is at index the variable
\code{x} holds.

It is implemented using \code{eval(substitute(sel), nl, parent.frame())}. The
first argument returns the expression passed to \code{sel}. The second is a
named list that contains the names of the columns from \code{df} and which will
be used to evaluate the expression. The last parameter indicates which
environment should be used to look up bindings not found in \c{nl}.

\paragraph{Boilerplate Removal} The combination of \code{match.call} and \code{eval}
is frequently used by modeling functions to avoid boilerplate code and ease
maintenance. Consider the example shown below of function \code{survfit.formula}
from the \code{survival} package.

\begin{figure}[h]
\begin{lstlisting}
survfit.formula <- function(formula, data, weights, ...) {
  Call <- match.call()
  indx <- match(c('formula', 'data', 'weights'), names(Call), nomatch=0)
  temp <- Call[c(1, indx)]
  temp[[1L]] <- quote(stats::model.frame)
  mf <- eval.parent(temp)
}
\end{lstlisting}
  \caption{survival::survfit.formula}
\end{figure}

Here, \code{match.call} is used to access the expression with which
\code{survfit.formula} is called. The next three lines construct a call to
\code{stats::model.frame} with argument text of parameters \code{'formula'},
\code{'data'}, and \code{'weights'}. Then, the call expression is evaluated in
the parent environment. The use of \code{match.call}, avoids repeated use of
\code{substitute} for extracting the argument expressions. This setup is also
extensible as new parameter names can be easily added to the \code{match} call.

\paragraph{Formula Resolution} Many packages use \code{eval} to evaluate formula
expressions. Consider the definition of the \code{formula.pdMat} function from
\code{nlme} package.

\begin{figure}[h]
\begin{lstlisting}
formula.pdMat <- function(x, asList, ...) eval(attr(x, "formula"))
\end{lstlisting}
  \caption{nlme::formula.pdMat}
\end{figure}

It extracts the \code{"formula"} attribute attached to the \code{pdMat} object
and evaluates it.

\paragraph{Code Obfuscation} Some packages use \eval to bypass the restrictions
imposed by the \code{R CMD CHECK} tool. This tool enforces some well-formedness
rules on a submitted package before accepting it for inclusion in CRAN. Some
static checks ensure that the package code does not use certain restricted
functions. If package authors cannot avoid using these functions, they use \eval
to obfuscate the code, bypassing the checks performed by the tool. An example of
this use is the \code{scalaEnsure} function from the \code{aibd} package.
\begin{figure}[h]
\begin{lstlisting}
scalaEnsure <- function() {
    ...
    env <- parent.env(environment())
    eval(parse(text=paste0('unlockBinding("s", env)')))
    assign("s", s, envir=env)
    lockBinding("s", env)
}
\end{lstlisting}
  \caption{aibd::scalaEnsure}
\end{figure}

This function mutates the variable \code{s} in the package environment. However,
the package environment in R is locked by default, which prevents bindings from
being mutated. Environment can be unlocked using \code{unlockBinding}, but its
use is flagged by the well-formedness checks. To work around this limitation,
the author passes the expression to unlock the package environment as a string
to \code{parse} which is then evaluated by \code{eval}.

\subsection{Shapes redux}

@@

    Another example is the most often called \eval site in CRAN:
    \lstinline|eval(`_inherit`, env)| defined in the \c{find\_super} function in
    the \c{ggproto} object system. It looks up the variable \c{`\_inherit`} to
    find the parent class. In the vast majority of cases, the variable is a
    symbol, so the \eval seems superfluous. The \eval is however necessary so
    one can specify parent classes using \c{pkg\_name::class\_name}, as the
    \c{::} operator is a function call in R.

@@

    There are also more sophisticated cases. The \c{plyr::allocate\_column}
    function uses \eval for a deferred assignment to fill missing values in a
    column. Package \c{overture} repeatedly evaluates R expressions capturing
    any assignments that happen in it as samples of Markov chains.

@@

    Call reflection refers to the combination of \eval with \c{match.call}, a
    function that returns the current call with all its arguments captured as
    expressions. This is used mostly in the following circumstances: (1) to
    record a call for later use in a possibly different environment, (2) pass
    most of the captured call arguments to another function, or (3) evaluate
    arguments of the function in different environments. This is widely used in
    relation to statistical modeling in combination with a model fitting
    function or with \c{model.frame} and related functions. For example the
    \c{survival::coxph} function\footnote{Function fitting a Cox proportional
    hazards regression model.} contains:
    %
    \begin{lstlisting}[gobble=2]
    Call <- match.call()
    tform <- Call[c(1,indx)]  # only keep the arguments we wanted
    tform[[1L]] <- quote(stats::model.frame)  # change the function called
    mf <- eval(tform, parent.frame())
    \end{lstlisting}

>>>>

The R language was intended to be extensible. The combination of lazy
evaluation, \c{substitute} and \c{eval} are some of the tools given to
developers to this end. To understand \emph{why} developers use \eval, we
manually inspected top 117 sites which contribute to 90\% of the calls in the
corpus. This section gives some real-world examples, we simplify code for
clarity.

\paragraph{Extended Interfaces}
Sometimes it is convenient to allow users to extend the behavior of a library
function. \Eval allows clients to provide a simple text string that shall be
executed in the context of the function. The example here allows users to
provide formulas as strings, such as \k{"exp(-x^2/2)"}, the contract being that
the text can refer to some variable \c x that will be set up by the called
function. In this case one users could provide an anonymous function.

\begin{figure}[h!]
\begin{lstlisting}
 function(n,formula) {
    p <- function(x) eval(parse(text=formula))
\end{lstlisting} \caption{AdapSamp::rARS}
\end{figure}

\paragraph{Code Generation}
\Eval can be used to generate boiler plate code that would either be tedious to
write by hand or that is only known after deployment. The example uses \eval to
construct function objects that are parameterized with string \c{what}.
This is used  automatically generate R bindings to native code.

\begin{figure}[h]
\begin{lstlisting}
 function(methods,env) {
   for(what in cppMethods)
     methods[[what]] <- eval(substitute(function(...)CppObject$WHAT(...,name=what)),env)
\end{lstlisting}
  \caption{Rcpp::.makeCppMethods}
\end{figure}

\paragraph{Extracting Varargs}
A somewhat odd use of \eval is to obtain the source code of the arguments passed
to a vararg parameter. In this example, the function iterates over the contents
of the vararg until it finds a non-null value. \k{substitute(alist(...))} yields
an expression that represents a call to \k{alist} but where the \code{...} are
expended with the arguments supplied to the function. \Eval then returns a
\code{list} with unevaluated argument expressions which are evaluated
sequentially for their result.

\begin{figure}[h]
\begin{lstlisting}
 function(...) {
   for(e in eval(substitute(alist(...)))) {
     x <- eval(e, parent.frame())
     if(!is.null(x)) break
   }
   x
 }
\end{lstlisting}
  \caption{statnet.common::NVL}
\end{figure}

\paragraph{Controlled Evaluation}
Developer often want to control when and if expressions are evaluated. For
example, a package that provides users with the ability to make assertions and,
if they fail, produce legible error messages. The example here accepts a vararg
with a list of expressions whose code is extracted using
\code{eval(substitute(alist(...)))}. The expression are evaluated sequentially
in the caller's environment. If an expression is false, an error message is
generated.

\begin{figure}[h]
\begin{lstlisting}
 function(..., env=parent.frame(), msg=NULL) {
    asserts <- eval(substitute(alist(...)))
    for (assertion in asserts) {
      res <- tryCatch({
        eval(assertion, env)
      }, assertError=function(e) { structure(FALSE, msg=e$message) })
\end{lstlisting}
  \caption{assertthat::see\_if}
\end{figure}

\paragraph{Shadowing}
%%% TODO ?????????????////
Many packages use \eval to implement variations of the \code{base} R functions.
For example, the \code{igraph} package provides the \code{do_call} function
shown below. This is a variant of the \code{do.call} function from the
\code{base} package. This function executes a function call by constructs it
from a name and a list of arguments and calling \eval.

\begin{figure}[h]
\begin{lstlisting}
 function(f,...,args=list(),env=parent.frame()) {
   f <- substitute(f)
   eval(make_call(f,...,args), env)
 }
\end{lstlisting}
\caption{igraph::do\_call}
\end{figure}

\paragraph{Domain-Specific Language}
A domain-specific language leverages R's grammar but redefines it semantics in
some suitable way. The usage example illustrates string interpolation. The
\k{glue} function extracts snippets of code enclosed between braces and
evaluates them using \eval, and finally splices their results to construct the
result.

@@

\paragraph{Boilerplate Removal}
A variant of code generation is to let users write compact code that is expended
via \eval and \code{match.call}. The example uses \code{match.call} to access
the expression with which the function is called. The next three lines construct
a call to \code{stats::model.frame} with argument text of parameters
\code{'formula'}, \code{'data'}, and \code{'weights'}. Then, the call expression
is evaluated in the parent environment. The use of \code{match.call}, avoids
repeated use of \code{substitute} for extracting the argument expressions. This
setup is also extensible as new parameter names can be easily added.

\begin{figure}[h]
\begin{lstlisting}
 function(formula, data, weights, ...) {
   Call <- match.call()
   indx <- match(c('formula', 'data', 'weights'), names(Call), nomatch=0)
   temp <- Call[c(1, indx)]
   temp[[1L]] <- quote(stats::model.frame)
   mf <- eval.parent(temp)
\end{lstlisting}
  \caption{survival::survfit.formula}
\end{figure}

\paragraph{Formula Resolution}
Many packages use \eval to evaluate formulas. The example extracts the
\code{"formula"} attribute attached to the \code{pdMat} object and evaluates it.

\begin{figure}[h]
\begin{lstlisting}
 function(x, asList, ...) eval(attr(x, "formula"))
\end{lstlisting}
  \caption{nlme::formula.pdMat}
\end{figure}

\paragraph{Obfuscation}
Some packages use \eval to bypass the restrictions imposed by the \code{R CMD
  CHECK} tool. This tool enforces some well-formedness rules on a submitted
package before accepting it for inclusion in CRAN. Some static checks ensure
that the package code does not use certain restricted functions. Package authors
use \eval to obfuscate their code. The example mutates the variable \code{s} in
the package environment. However, that is locked by default and
\code{unlockBinding}, is a restricted function. To work around this, the call to
unlock is done through \eval.

\begin{figure}[h]
\begin{lstlisting}
 function() {
   env <- parent.env(environment())
   eval(parse(text=paste0('unlockBinding("s", env)')))
   assign("s", s, envir=env)
   lockBinding("s", env)
\end{lstlisting}
  \caption{aibd::scalaEnsure}
\end{figure}

@@

@@

@@
